Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-10-22 notebooks/expected_and_actual_counts_ratio.Rmd
Calculate fold change and create volcano plot

@@

```{r contingency}
# CREATE SUBSET OF DATA TO WORK WITH: ONLY CLADES 20A AND 20B FOR ALL COUNTRIES
temp <- filtered_SARS_mutation
filtered_SARS_mutation <- temp[which(temp$clade=="20A" | temp$clade=="20B"),]

# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, filtered_SARS_mutation$clade, sep="_"), observed=filtered_SARS_mutation$actual_count, expected=filtered_SARS_mutation$expected_count)

# Identify unique clades in the dataset.
clades <- unique(filtered_SARS_mutation$clade)

# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)

# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
nt_mutation_list <- unique(paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, sep="_"))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))

# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]

# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))

@@

# Combine p-value list with the reference list of clade comparisons.

# run fishers exact an save p-value to dataframe (clade comparison, site, mutation, p-value, fdr p-value, bonferroni p-value)

# test run volcano plot for each clade comparison

>>>>

#Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count+0.5)/(filtered_SARS_mutation$expected_count+0.5)

@@

```{r contingency}
#CREATE SUBSET OF DATA TO WORK WITH: ONLY CLADES 20A AND 20B FOR ALL COUNTRIES
all_data <- filtered_SARS_mutation
filtered_SARS_mutation <- all_data[which(all_data$clade=="20A" | all_data$clade=="20B"),]

#Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, filtered_SARS_mutation$clade, sep="_"), observed=filtered_SARS_mutation$actual_count, expected=filtered_SARS_mutation$expected_count, observed_expected=filtered_SARS_mutation$observed_expected)

#Identify unique clades in the dataset.
clades <- unique(filtered_SARS_mutation$clade)

#Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)

#Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
nt_mutation_list <- unique(paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, sep="_"))

#In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))

#In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

#Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))

#Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]

#Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))

@@

#Combine p-value list with the clade comparison list.
comparison_p_values <- cbind(filtered_comparisons, p_values)

#False Discovery Rate and Bonferroni Corrections

#Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]),"observed_expected"],fisher_data[which(fisher_data$nt_mutation_clade==c[2]),"observed_expected"])})

#Volcano Plot
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point()
